---
title: "results"
output: html_document
date: "2024-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(papaja)
require(Hmisc)
require(forcats)

d = read_csv('data/Decoding Development Data DStaton Research (Responses) - Grid Data.csv') %>% 
  left_join(read_csv('data/Decoding Development Data DStaton Research (Responses) - Participant Identifying # and basic information.csv'), by = join_by(participant_id == `Identifying Number`)) %>% 
  left_join(
read_csv('data/Decoding Development Data DStaton Research (Responses) - Assessment data.csv') %>% 
  select(participant_id = `Identifying Number`,
         ctopp_elision = `CTOPP - elision`,
         ctopp_letter_naming = `CTOPP - rapid letter naming`,
         wj4 = WJ4)) %>% 
  mutate(x = as.numeric(x),
         y = as.numeric(y)) %>% 
  filter(!is.na(x) | !is.na(y)) %>% 
  glimpse()

assessments = read_csv('data/Decoding Development Data DStaton Research (Responses) - Assessment data.csv') %>% 
  select(participant_id = `Identifying Number`,
         ctopp_elision = `CTOPP - elision`,
         ctopp_letter_naming = `CTOPP - rapid letter naming`,
         wj4 = WJ4, grade = Grade)

source("model_predictions.R")

pca = d %>% 
  select(x, y) %>% 
  prcomp()

d = cbind(d, pca$x)

dms = list()

for (id in unique(d$participant_id)){
  
  df = d %>% 
    filter(participant_id == id) %>% 
    select(letter, x, y)
  
  dm = df %>% 
    select(x, y) %>% 
    dist()
  
  tmp = as.matrix(dm)
  rownames(tmp) = df$letter
  colnames(tmp) = df$letter
  
  dms[[id]] = as.data.frame(as.table(tmp)) %>% 
    rename(letter_1 = Var1, letter_2 = Var2, distance = Freq) %>% 
    filter(letter_1 != letter_2) %>% 
    filter(as.numeric(factor(letter_1)) > as.numeric(factor(letter_2))) %>% 
    mutate(participant_id = id)
  
}

vowels = c("a", "e", "i", "o", "u", "y")

distances = purrr::list_rbind(dms) %>% 
  mutate(letter_1_type = case_when(letter_1 %in% vowels ~ "vowel",
                          TRUE ~ "consonant"),
         letter_2_type = case_when(letter_2 %in% vowels ~ "vowel",
                          TRUE ~ "consonant")) %>% 
  left_join(assessments) %>% 
  mutate(order = case_when(grade == "K" ~ 1,
                           grade == "3" ~ 2))

MSD = distances %>% 
  group_by(participant_id) %>% 
  summarise(mean_distance = mean(distance, na.rm = T),
            sd_distance = sd(distance, na.rm = T))

distances = distances %>% 
  left_join(MSD) %>% 
  mutate(distance_z = (distance - mean_distance)/ sd_distance) %>% 
  left_join(letter_distances, by = c("letter_1", "letter_2"))


rm(df, dm, dms, tmp, letter_distances)
```


## Distances between target letters

### "f" and "v"
These are the average distances between the letters "f" and "v" among participants. The errorbars are standard errors of the mean.

```{r}
N = length(unique(distances$participant_id))

plot = distances %>% 
  filter(letter_1 == "v" & letter_2 == "f") %>% 
  group_by(grade) %>% 
  summarise(M = mean(distance, na.rm = T),
             SD = sd(distance, na.rm = T),
             SEM = SD/sqrt(N),
             grade = first(grade),
             order = first(order)) %>% 
  ggplot(aes(reorder(grade, order), M, fill = factor(grade))) +
  geom_bar(stat = "summary", color = "black") +
  geom_errorbar(aes(ymin = M - SEM, ymax = M + SEM), width = .2) +
  theme_apa() +
  labs(title = 'The distance between "f" and "v"',
       x = 'Grade', y = 'Distance') +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("skyblue4", "darkorange3"))

ggsave("figure_1.png", plot)
```


### "p" and "b"
```{r}


plot = distances %>% 
  filter(letter_1 == "p" & letter_2 == "b") %>% 
  group_by(grade) %>% 
  summarise(M = mean(distance, na.rm = T),
             SD = sd(distance, na.rm = T),
             SEM = SD/sqrt(N),
             grade = first(grade),
             order = first(order)) %>% 
  ggplot(aes(reorder(grade, order), M, fill = factor(grade))) +
  geom_bar(stat = "summary", color = "black") +
  geom_errorbar(aes(ymin = M - SEM, ymax = M + SEM), width = .2) +
  theme_apa() +
  labs(title = 'The distance between "p" and "b"',
       x = 'Grade', y = 'Distance') +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("skyblue4", "darkorange3"))

ggsave("figure_2.png", plot)
```


## Correlation between orthographic distance and placement distance
Here we focus on letters which are associated with specific/ reliable phonemes. We would expect for these letters that less skilled children would be more influenced by their visual properties.

### Vowels
```{r}

targets = c("b", "d", "f", "g", "h", "j", "k", "l", "m", "n", "p", "r", "s", "t", "v", "w", "z", "c", "q", "x")

annotations = distances %>% 
  filter(letter_1 %nin% targets & letter_2 %nin% targets) %>% 
  group_by(grade) %>% 
  summarise(corr = cor(orthographic_distance_z, distance_z, method = "spearman")) %>%
  mutate(annotation = str_c("ρ = ", round(corr, digits = 2)),
         grade = factor(grade, levels = c("K", "3")),
         x = c(-1.0, -1.0),
         y = c(2, 2))


plot = distances %>% 
  filter(letter_1 %nin% targets & letter_2 %nin% targets) %>% 
  mutate(grade = factor(grade, levels = c("K", "3"))) %>% 
  ggplot(aes(orthographic_distance_z, distance_z, color = grade)) +
  #geom_point() +
  geom_jitter(size = .5, width = .02) +
  geom_smooth(method = "lm") +
  facet_wrap(~grade) +
  theme_apa() +
  labs(title = "Vowels", x = "Orthographic distance (from ANN)", y = "Student rated distance") +
  theme(legend.position = "none") +
  scale_color_manual(values = c("darkorange3", "skyblue4")) +
  geom_text(data = annotations, aes(x = x, y = y, label = annotation), size = 6, color = "black")

plot


ggsave("figure_3a.png", plot)




```

### Consonants
```{r}

targets = c("b", "d", "f", "g", "h", "j", "k", "l", "m", "n", "p", "r", "s", "t", "v", "w", "z", "c", "q", "x")


annotations = distances %>% 
  filter(letter_1 %in% targets & letter_2 %in% targets) %>% 
  group_by(grade) %>% 
  summarise(corr = cor(orthographic_distance_z, distance_z, method = "spearman")) %>%
  mutate(annotation = str_c("ρ = ", round(corr, digits = 2)),
         grade = factor(grade, levels = c("K", "3")),
         x = c(1.8, 1.8),
         y = c(2.7, 2.7))

plot = distances %>% 
  filter(letter_1 %in% targets & letter_2 %in% targets) %>% 
  mutate(grade = factor(grade, levels = c("K", "3"))) %>% 
  ggplot(aes(orthographic_distance_z, distance_z, color = grade)) +
  #geom_point() +
  geom_jitter(size = .5, width = .02) +
  geom_smooth(method = "lm") +
  facet_wrap(~grade) +
  theme_apa() +
  labs(title = "Consonants", x = "Orthographic distance (from ANN)", y = "Student rated distance") +
  theme(legend.position = "none") +
  scale_color_manual(values = c("darkorange3", "skyblue4")) +
  geom_text(data = annotations, aes(x = x, y = y, label = annotation), size = 6, color = "black")

plot
ggsave("figure_3b.png", plot)


```


## Relative similarity of vowels versus consonants

```{r}
vowels = c("a", "e", "i", "o", "u", "y")
consonants = letters[letters %nin% vowels]

distances %>% 
  mutate(condition = case_when((letter_1 %in% vowels & letter_2 %in% vowels) ~ "vowel",
                               (letter_1 %in% consonants & letter_2 %in% consonants) ~ "consonant",
                               TRUE ~ "mixed")) %>% 
  ggplot(aes(condition, distance_z, fill = condition)) +
  geom_bar(stat = "summary", color = "black") +
  facet_wrap(~grade) +
  theme_apa() +
  labs(fill = "Condition", x = "Condition", y = "Distance (standardized)") +
  scale_fill_manual(values = c("pink", "purple", "turquoise"))
  

```




## As a function of skill level

### Figure 4
```{r}

distance_correlation_per_participant = tibble(participant_id = unique(distances$participant_id))

cors = c()

for (participant in unique(distances$participant_id)){
  
  
  rho  = distances %>% 
    filter(participant_id == participant) %>% 
    summarise(cor(distance_z, orthographic_distance_z, method = "spearman"))
  
  cors = c(cors, rho$`cor(distance_z, orthographic_distance_z, method = "spearman")`)
  
}

distance_correlation_per_participant$rho = cors

distance_correlations = distances %>% 
  group_by(participant_id) %>% 
  summarise(wj4 = first(wj4),
            ctopp_elision = first(ctopp_elision),
            ctopp_letter_naming = first(ctopp_letter_naming),
            grade = first(grade)) %>% 
  left_join(distance_correlation_per_participant)


Ms = distance_correlations %>% 
  group_by(grade) %>% 
  summarise(rho = mean(rho)) %>% 
  mutate(grade = forcats::fct_relevel(grade, "K", "1"))


annotations = distance_correlations %>% 
  filter(letter_1 %in% targets & letter_2 %in% targets) %>% 
  group_by(grade) %>% 
  summarise(corr = cor(orthographic_distance_z, distance_z, method = "spearman")) %>%
  mutate(annotation = str_c("ρ = ", round(corr, digits = 2)),
         grade = factor(grade, levels = c("K", "3")),
         x = c(1.8, 1.8),
         y = c(2.7, 2.7))


annotations = distance_correlations %>% 
  arrange(-desc(ctopp_elision)) %>% 
  mutate(rank = seq_len(n()),
         order = case_when(grade == "K" ~ 1,
                           grade == "1" ~ 1)) %>% 
  group_by(grade) %>% 
  summarise(corr = cor(rank, rho)) %>% 
  mutate(annotation = str_c("ρ = ", format(round(corr, digits = 2))),
         grade = factor(grade, levels = c("K", "3")),
         x = c(12, 8),
         y = c(.09, .1))

plot = distance_correlations %>% 
  arrange(-desc(ctopp_elision)) %>% 
  mutate(rank = seq_len(n()),
         grade = fct_relevel(grade, "K", "1")) %>% 
  ggplot(aes(rank, rho, label = participant_id, fill = grade, color = grade)) +
  #ggplot(aes(ctopp_elision, rho, label = participant_id)) +
  geom_hline(data = Ms, linetype = "dashed", aes(yintercept = rho), color = "grey36") +
  geom_smooth(method = "lm") +
  geom_point(color = "black") +
  theme_apa() +
  labs(x = "Phonological processing", y = "Orthographic influence on rated distance", fill = "Grade") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("darkorange3", "skyblue4")) +
  scale_color_manual(values = c("darkorange3", "skyblue4"))  +
  facet_wrap(~grade, scales = "free_x") +
  geom_text(data = annotations, aes(x = x, y = y, label = annotation), size = 6, color = "black")

plot


ggsave("png/figure_4a.png", plot)



annotations = distance_correlations %>% 
  mutate(order = case_when(grade == "K" ~ 1,
                           grade == "1" ~ 1)) %>% 
  group_by(grade) %>% 
  summarise(corr = cor(wj4, rho)) %>% 
  mutate(annotation = str_c("ρ = ", format(round(corr, digits = 2))),
         grade = factor(grade, levels = c("K", "3")),
         x = c(35, 25),
         y = c(.1, .1))


plot = distance_correlations  %>% 
  mutate(rank = seq_len(n()),
  grade = fct_relevel(grade, "K", "1")) %>% 
  #ggplot(aes(wj4, rho, label = participant_id, fill = grade)) +
  ggplot(aes(wj4, rho, label = participant_id, fill = grade, color = grade)) +
  geom_smooth(method = "lm") +
  #geom_smooth(method = "loess", se = F, linetype = "dashed", color = "black") +
  geom_point(color = "black") +
  theme_apa() +
  facet_wrap(~grade, scales = "free_x") +
  geom_hline(data = Ms, linetype = "dashed", aes(yintercept = rho), color = "grey36") +
  labs(x = "Letter/word ID", y = "Orthographic influence on rated distance") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c( "darkorange3", "skyblue4")) +
  scale_color_manual(values = c("darkorange3", "skyblue4")) +
  geom_text(data = annotations, aes(x = x, y = y, label = annotation), size = 6, color = "black")

plot

ggsave("png/figure_4b.png", plot)





```


